Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-08-29 paper.tex
web-like -> weblike, spider-web -> spiderweb

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic
reconstruction from limited data.  The classical descriptors are 2-point
correlation functions (CFs), but the computation of 3-point statistics is known
to be advantageous in some cases as they can probe non-Gaussian signatures, not
captured by their 2-point counterparts. Moreover, $n$-point CFs with $n \ge 3$
are believed to possess larger information content and provide more information
about studied structures. In this paper, we have developed algorithms and code
to compute $S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle
triangle pattern. This pattern is much easier to vectorize and parallelize
compared to previously used random triangle or spider-web sampling
patterns. Moreover, it allows to easily implement computations of directional
3-point CFs. The execution times for the same number of samples are orders of
magnitude lower than in existing published counterparts. We showed that the volume
of data produced get unwieldy very easily, especially if computations are
performed in frequency domain. For these reasons until information content of
different sets of correlation functions with different "n-pointness" is not
known, advantages of CFs with $n>3$ are not clear. Nonetheless, developed
algorithms and code are universal enough to be easily extendable to any $n$
(with increasing computational and RAM burden). All results are available as
part of open source package \code{CorrelationFunctions.jl} [Postnicov et al.,
  2024, Comp.Phys.Commun., 299, 109134]. As described in this paper, 3-point CFs
computations can be immediately applied in a great number of research
applications, for example: 1) flow and transport velocity fields analysis or any
data with non-Gaussian signatures, 2) deep learning for structural and physical
properties, 3) structure taxonomy and categorization.  In all these and numerous
other potential cases the ability to compute directional 3-point functions may
be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images
(while binary structures were used in this paper for simplicity of
explanations).
\end{abstract}

@@

We have made execuation time comparisons of our algorithm with existing
algorithms.  However, it is also beneficial to compare some algorithmic details
against existing digital approaches -- proposed by Malmir~et~al
\cite{malmir2018} and another one by Smith and Torquato \cite{SMITH1988176}. The
first algorithm generates a triangle with fixed sides $r_1$ and $r_2$ and an
arbitrary angle $\theta$ between them and places it in the medium firstly making
a random translation and a rotation around a random vector by a random angle. By
making a large number of trials the algorithm gets a value of a three-point
correlation function at the point $(r_1, r_2, \theta)$. The second algorithm
takes a precalculated triangle $(r_1, r_2, \theta)$ from a web-like pattern and
places it in the medium without rotation, only applying a random
translation. This way the algorithm preserves orientation of triangles stored in
the pattern. The main advantage of our approach based on right triangle is in
its parallelizability and ease of implementation. A drawback is that an angle
between catheti of the sampling pattern is fixed to be $\theta = \frac{\pi}{2}$,
but it can be mitigated by choosing a random pattern or rotating the
pattern. The rotations can be random as in the Malmir et al. algorithm, or it
can be a single rotation to align catheti of the sampling triangle with the
direction of anisotropy (\cref{fig:aniso}). The latter allows computing
directional 3-point correlation functions for anisotropic structures similar to
2-point counterparts \cite{10.1063/1.4867611,EPL1}.

>>>>

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic
reconstruction from limited data.  The classical descriptors are 2-point
correlation functions (CFs), but the computation of 3-point statistics is known
to be advantageous in some cases as they can probe non-Gaussian signatures, not
captured by their 2-point counterparts. Moreover, $n$-point CFs with $n \ge 3$
are believed to possess larger information content and provide more information
about studied structures. In this paper, we have developed algorithms and code
to compute $S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle
triangle pattern. This pattern is much easier to vectorize and parallelize
compared to previously used random triangle or spiderweb sampling
patterns. Moreover, it allows to easily implement computations of directional
3-point CFs. The execution times for the same number of samples are orders of
magnitude lower than in existing published counterparts. We showed that the volume
of data produced get unwieldy very easily, especially if computations are
performed in frequency domain. For these reasons until information content of
different sets of correlation functions with different "n-pointness" is not
known, advantages of CFs with $n>3$ are not clear. Nonetheless, developed
algorithms and code are universal enough to be easily extendable to any $n$
(with increasing computational and RAM burden). All results are available as
part of open source package \code{CorrelationFunctions.jl} [Postnicov et al.,
  2024, Comp.Phys.Commun., 299, 109134]. As described in this paper, 3-point CFs
computations can be immediately applied in a great number of research
applications, for example: 1) flow and transport velocity fields analysis or any
data with non-Gaussian signatures, 2) deep learning for structural and physical
properties, 3) structure taxonomy and categorization.  In all these and numerous
other potential cases the ability to compute directional 3-point functions may
be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images
(while binary structures were used in this paper for simplicity of
explanations).
\end{abstract}

@@

We have made execuation time comparisons of our algorithm with existing
algorithms.  However, it is also beneficial to compare some algorithmic details
against existing digital approaches -- proposed by Malmir~et~al
\cite{malmir2018} and another one by Smith and Torquato \cite{SMITH1988176}. The
first algorithm generates a triangle with fixed sides $r_1$ and $r_2$ and an
arbitrary angle $\theta$ between them and places it in the medium firstly making
a random translation and a rotation around a random vector by a random angle. By
making a large number of trials the algorithm gets a value of a three-point
correlation function at the point $(r_1, r_2, \theta)$. The second algorithm
takes a precalculated triangle $(r_1, r_2, \theta)$ from a weblike pattern and
places it in the medium without rotation, only applying a random
translation. This way the algorithm preserves orientation of triangles stored in
the pattern. The main advantage of our approach based on right triangle is in
its parallelizability and ease of implementation. A drawback is that an angle
between catheti of the sampling pattern is fixed to be $\theta = \frac{\pi}{2}$,
but it can be mitigated by choosing a random pattern or rotating the
pattern. The rotations can be random as in the Malmir et al. algorithm, or it
can be a single rotation to align catheti of the sampling triangle with the
direction of anisotropy (\cref{fig:aniso}). The latter allows computing
directional 3-point correlation functions for anisotropic structures similar to
2-point counterparts \cite{10.1063/1.4867611,EPL1}.
